---
title: "Bivariate Choropleth Plot using sf and ggplot package"
author: Fahim Ahmed
date: '2020-11-18'
slug: bivariate-choropleth-plot-using-sf-and-ggplot-package
categories:
  - map
  - choropleth
tags:
  - R
  - ggplot2
  - sf
  - map
  - choropleth
editor_options: 
  chunk_output_type: console
---

```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      include=TRUE, 
                      message=FALSE, 
                      warning=FALSE)

```

## Goal

The goal of this blogpost is to learn how to draw a bivariate map
using "ggplot2" and "sf" package.

I am very closely follwing the instruction from this [tutorial](https://www.trafforddatalab.io/recipes/gis/bivariate_choropleth.html)

## A little about the dataset

The example dataset contains  a list for the geophaphy of the United Kingdom (UK) at UTLA level and related attribute table. You can think the UTLA as a small geographic unit (i.e., region) in the UK. These geographic information are stored in the sf object.  In the attribute table, the covid rate (tested +ve per 1000 people) and population (%) by ethnicity is collected from the ONS database at each UTLA level. 


## Question/Objective:

The question related to the dataset is: *How the covid rate changes by ethnic population (%) at UTLA level? *

So, the objective is to map, 

- covid rate (x variable)
- ethnic population (y variable)
- at UTLA (geophaphic variable)

One of the tools that can be used to answer the question is using a bivariate Choloropleth map.


## Step-by-step procedure


## Assumption

We already created a dataset of type "list". This dataset contains a "sf" object for geographic information and a dataframe provides a attributes related to geography.

1. Load Necessary Libaries and Datasets

```{r}
library(here)
library(janitor) # tallying categorical variable
library(cowplot)  # adding legend over the ggplot map
library(tidyverse) # 
library(sf) # read/write/wrangle geographic data
library(classInt) # creating discrete intervals from continuous var
library(biscale)
library(pals) # color palette and color scale
```


```{r}
df <- st_read(here("data_geography", "uk_utla_covid", "uk_utla_covid.shp"))
```

2. We want to add a $bi_class$ variable to the attribute dataset. The $bi\ class$ variable incorporates the change in $x$ variable and $y$ variable. We can create a $2x2$ or $3x3$ bivariate $bi class$ variables. Joshua Stevens [blog post](https://www.joshuastevens.net/cartography/make-a-bivariate-choropleth-map/) provides detail, step by step guide on bivariate plot.  The idea is that we want to divide the (continuous) values of $x$ variable into $2$ or $3$ segment. We will do the same division to the $y$ variable. For example, the varible $covid\ rate$classIntervals(df_clean$covid_rate, n=3, style="quantile") is divided into 3 parts using "quantile" division method. 

```{r}
classInt::classIntervals(df$covd_rt, n=3, style="quantile")
```

The three division quatiles values for covid rate are: 

- $1:[6.205,16.08)$
- $2:[16.08,27.22)$
- $3:[27.22,60.46]$

The $1,2,3$ here represent the classes for covid rate. A vizualization for the quantile division is presented below:

```{r}
pal1 <- c("wheat1", "red3")
plot(classInt::classIntervals(df$covd_rt, n=3, style="quantile"), pal=pal1, main="Quantile")
```

Similary, the 3 divisions for the black population is below:
```{r}
classInt::classIntervals(df$black, n=3, style="quantile")
```

The three classes for the population of black ethnic group are:

- 1: [0.09025,0.6808)
- 2: [0.6808,2.802)
- 3: [2.802,27.16] 


```{r}
plot(classInt::classIntervals(df$black, n=3, style="quantile"), pal=pal1, main="Quantile")
```

The following code creates the $bi\ class$ variable. It maps the x (=covid rate) division and y divisions (=black pop.) together.

```{r}
## A "df_clean" list is already created that contains a "sf" object and
## a "dataframe"


## Create a new column "bi_class" using x, y variable that
## 3 by 3; quantile division of the variables

data <-
  bi_class(
    df,     # <--- INPUT data set
    x = covd_rt,  # <-- variable 1 INPUT
    y = black,  # <-- variable 2 INPUT
    style = 'quantile',
    dim = 3
    )


```

The first six observations of variable $covid\ rate$, $black\ population$, and newly created $bi\ class$ from the dataset is presented below:

```{r, include=FALSE}
data %>% dplyr::select(covd_rt, black, bi_class) %>% head()
```

The above table indicates that the values for first observation for covid rate and black poulation is $37.38$ and $0.185$, respectively. It indicates that the covid rate is between $3:[27.22,60.46]$ values and black population is between $[0.09025,0.6808)$ values.  These quantile bins were created previously.  Thus the bi class variable considers both values from both variable and assigns a class for each observation; the value for the first observation is $3-1$.

Before diving into the map plotting, the number of UTLA regions associated with each $bi\ class$ variable are shown below. The nine classes are shown in y-axis and the number of UTLA regions associated with it are shown. The class 2-3 has the highest number of UTLA regions (=22). The class 1-2 and 3-1 both have equal number of regions (i.e., 21).

```{r}
data %>% 
  tabyl(bi_class) %>%
  ggplot()+
  geom_col(aes(x = n,
               y = bi_classgeometry), 
           width = 0.3) +
  scale_x_continuous(breaks = seq(0,25, by = 1)) +
  theme_bw() +
  labs(x = "Number of UTLA regions",
       y = "Class Asssigned for COVID-19 rate and Black Pop. (%)")
```

In the next step, the $bi\ class$ variable is mapped using ggplot and sf package.  


```{r}
## Plotting
map <- ggplot() +
  geom_sf(
    data = data,
    mapping = aes(fill = bi_class),
    color = "white",
    size = 0.1,
    show.legend = FALSE
  ) +
  bi_scale_fill(pal = "DkViolet", dim = 3) +
  labs(title = "",
       caption = "Source: 2011 Census | Sport England") +
  coord_sf(crs = st_crs(4326), datum = NA) +
  theme_void() +
  theme(
    plot.margin = unit(c(1, 1, 1, 1), "cm"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.caption = element_text(
      size = 9,
      color = "grey50",
      hjust = 1,
      margin = margin(t = 15)
    )
  )

```

In additon, the legend for bivariate is created separately for the map. 

```{r}
# create legend
# options for color for bi_class mapping: pal	one of "Brown", "DkBlue", "DkCyan", "DkViolet", or "GrPink".

legend <-
  bi_legend(
    pal = "DkViolet",
    dim = 3,
    xlab = "Higher covid rate",
    ylab = "Higher black pop. (%)",
    size = 8
  )
```

In this step, the map and the legend are combined to develop the final map.

```{r}

## combine map and legend

plot <- ggdraw() +
  draw_plot(map, x = 0, y = 0, width = 1, height = 1, scale = 1) +
  draw_plot(legend, x = 0.6, y = .6, width = 0.25, height = 0.25)
```

The final map is shown below:

```{r}
plot
```


<!-- ```{r} -->
<!-- ## save the plot -->

<!-- save_plot( -->
<!--   "bivariate_choropleth.png", -->
<!--   plot, -->
<!--   base_height = 8, -->
<!--   base_aspect_ratio = 1.8 -->
<!-- ) -->
<!-- ``` -->




