---
title: 'TT01:  Transit Cost Projects'
author: ''
date: '2021-01-09'
slug: tt01-transit-cost-projects
categories:
  - tidytuesday
tags:
  - R
  - visualization
  - ggplot2
  - tidytuesday
editor_options: 
  chunk_output_type: console
---


```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      include=FALSE, 
                      message=FALSE, 
                      warning=FALSE)

```

```{r}
# load packages
library(naniar)
library(ggplot2)
library(tidyverse)
```


This week's tidytuesday dataset explores the cost of transit projects for different countries of the world. The data is downloaded from the github repository.

```{r include=TRUE}
# read in the data manually

transit_cost <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-01-05/transit_cost.csv')



```

Let's look at the cost of transit projects in Japan (JP) and United States (US). Both of these countries are categorized as developed countries, but are located in different continent.  The data vizualization is focused on answering, **"How the transit projects' costs differ in two developed countries (JP vs US) of the world?"**


```{r}
transit_cost$start_year <- as.numeric(transit_cost$start_year)
transit_cost$end_year <- as.numeric(transit_cost$end_year)
transit_cost$rr <- as.factor(transit_cost$rr)
```
 

```{r include=TRUE}
# transit projects of us and japan

us_jp <- transit_cost %>% 
  filter(country %in% c("JP", "US"))  %>%
  mutate(railroad = ifelse(rr == 1, "RailRoad", "Not_RailRoad")) %>% 
  group_by(country) %>% 
  mutate(avg_cost_per_km = mean(cost_km_millions))

```

```{r}
# copied from Rpubs 
#Define gppr_theme() function

theme_gppr <- function(){ 
    font <- "Georgia"   #assign font family up front
    
    theme_bw() %+replace%    #replace elements we want to change
    
    theme(
      
      #grid elements
      #panel.grid.major = element_blank(),    #strip major gridlines
      panel.grid.minor = element_blank(),
      axis.ticks = element_blank(),          #strip axis ticks
      
      #since theme_minimal() already strips axis lines, 
      #we don't need to do that again
      
      #text elements
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 14,                #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   vjust = 2),               #raise slightly
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14),               #font size
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 9,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10),               #font size
      
      axis.text = element_text(              #axis text
                   family = font,            #axis famuly
                   size = 9),                #font size
      
      axis.text.x = element_text(            #margin for axis text
                    margin=margin(5, b = 10))
      
      #since the legend often requires manual tweaking 
      #based on plot content, don't define it here
    )
}
```

Both countries have almost similar number of projects (JP:15 vs. US: 13) in the dataset.  To compare the cost of transit projects, the "cost per kilometer" variable is chosen.  The cost per kilometer of each project is plotted against the start year of the project.  Also, whether the transit project was a railroad or not is included in the plot.


```{r include=TRUE}
us_jp %>%  ggplot() +
  # scatter plot of year vs cost per kilometer
  geom_point(
    aes(
      x = start_year,
      y = cost_km_millions,
      color = country
    ),
    size = 6
  ) +
  # adding name of the project for each data point
  geom_text(
    aes(x=start_year, y = cost_km_millions, label = line),
    nudge_x = 1,
    nudge_y = 2,
    angle = 90,
    check_overlap = T
  ) +
  # scaling x and y axis to accomodate the texts from names of projects
  scale_y_continuous(limit = c(0, 4500), expand = c(0 , 1000, 0, 500),  breaks = seq(0, 4500, by = 500), labels = scales::dollar_format()) +
  scale_x_continuous(limit = c(1985, 2025),
                     breaks = seq(1985, 2025, by = 5)) +
  # facetting the data for whether the project is railroad or not
  facet_wrap(vars(railroad)) +
  labs(
    title = "Comparison of transit projects' cost (per kilometer in USD) between Japan and USA",
    x = "Transit Project Start Year",
    y = "Cost per kilometer (USD)",
    color = "Country",
    shape = "Railroad Project?",
    caption = "Data Source: TidyTuesday 2021"
  ) + theme_gppr()
```

From the figure, it is quiet clear that Japan had more railroad transit projects than US.  Also, the average cost of the transit projects for Japan is lower that US.  The "East Side Access" railroad transit project was the cosliest one for US.  The transit projects of Japan started much earlier than US. Note that, this insights is only from using this particular dataset, and may not be generalized.