---
title: 'Classification Decision Tree in R using an Example Dataset '
author: ''
date: '2020-09-09'
slug: classification-decision-tree-in-r-using-an-example-dataset
categories:
  - machine learning
  - classification
  - R
tags:
  - R
  - rpart
  - decision tree
  - classification
draft: yes
---

# ML Classification Problem in R: Decision Tree



```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      include=FALSE, 
                      message=FALSE, 
                      warning=FALSE)

```

## Blogpost Highlights

### R Libraries used in Analysis

```{r}
library(rpart)
library(ggplot2)
library(skimr)
library(tidyverse)
library(tidymodels)
```

## Purpose 

The aim of this blog post is to learn:

+ detail calculation steps for a ML Classification method, namely Decision Tree 
+ how to use the tidymodels package for modeling 


## Assumptions

## Data Description

### Data Source

```{r}
penguins.csv <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-28/penguins.csv')
```

### Explore the data

```{r}
skim(penguins.csv)
```

The penguin dataset provides observation of 344 penguins and their characteristics . In the dataset, characteristics of three species of penguin are recorded over three years (i.e., 2007, 2008, and 2009).  The characteristic of penguin includes sex, island where it lives, bill length, bill depth, flipper length, and body mass.


## Big Question

Can you predict the penguin species using the features in the dataset?

## Methodology

### Build models

Step 1: Split dataset into training and testing datasets

The recipe packages from the tidymodels meta package is used to make training and testing datasets.  The penguin data set is split 3:1 ratio for training and testing datasets.  As a result, the training dataset contains `r nrow(penguin_train)` observations and testing dataset contains  `r nrow(penguin_test)` observations.

```{r}
# set the seed so that we have same dataset whenever we split the dataset
set.seed(1234)

# split the dataset
penguin_split <- initial_split(penguins.csv)

# save the spitted data into training and testing dataset
penguin_train <- training(penguin_split)
penguin_test <- testing(penguin_split)
```


### Evaluate models




###



## Result


## References
```{r}
citation("palmerpenguins")
```




